Comparing Machine Learning algorithms for classifying cognitive states from fMRI data

### 2012-11-11 article/article.tex
Proof-read discussion/conclusion

\section{Methods}
\label{sec:methods}
We now continue by discussing the methods applied in our research. The general approach taken for each of the algorithms discussed consists of the following steps:
\begin{enumerate}
  \item Preprocessing
  \item Dimensionality reduction
  \item Training
  \item Testing
\end{enumerate}
Preprocessing was done to transform the dataset into a format suitable for input into the classification algorithm. Dimensionality reduction was required to cope with the high dimensionality of the dataset and may consist of both feature extraction as well as feature selection. Finally, the approach taken to training and testing the classifier deals with preventing overfitting the classifier to training data and verifying accuracy for cross-subject classification. The choices made in each of these steps may affect classifier accuracy, so to obain a complete understanding of classification accuracies we have tried several methods at each step. In the following sections, these methods will be described in more detail.\\
The implementation of the algorithms was done using Python combined with modules tailored to mathematics (NumPy, SciPy) and machine learning (ScikitLearn). The original dataset was made available as a MatLab file and we used preprocessing in MatLab to create a number of files containing modified versions of this data, such as normalization and ROI grouping (see section~\ref{preprocessing}). The source code for the algorithms is available from the authors on request.

@@

\textbf{\emph{ROI mapping.}} ROI mapping is a form of feature selection, wherein only a subset of the N-dimensional input space is used. As described in the introduction, ROIs are anatomically defined brain areas classified as relevant by an expert in the field. We used 7 of the 25 defined ROIs, labeled  'CALC' 'LIPL' 'LT' 'LTRIA' 'LOPER' 'LIPS' and  'LDLPFC'. These ROIs have previously been found to be most relevant in discriminating between the cognitive states of looking at sentences or pictures \cite{wa:2003betw}. To further reduce the number of features, average voxel values within these regions are taken as new input features. As other higher-order features such as the standard deviation, kurtosis and skewness within ROIs did not increase predictability, these were not included in the subsequent analysis. The disadvantage of using ROIs is that they require \textit{a priori} knowledge. When this knowledge is inaccurate, classification accuracy may not be optimal.

@@

\textbf{\emph{Temporal dimensionality reduction.}} The BOLD response peaks at approximately 5 seconds after a stimulus is presented (see Fig. 1). This means that the first scans after a stimulus may not represent activity related to this stimulus and should not be used for classification. To examine which scans provide maximal class information, we plotted the average BOLD activity per scan for all trials of all subjects, given a trial type and ROI. As can be seen in Figure~\ref{fig:boldroi}, scans 10 to 20 provide maximal information for the first stimulus, while scans 23 to 33 provide maximal information for the second stimulus. The other scans were excluded.      
\begin{figure}
	\centering
	\includegraphics[width=90mm]{figures/scan_hemodynamic_BOLD_roi}
  	\caption{Haemodynamic response for ROI "CALC". The red line depicts a sentence-picture trial (S-P) and the blue line depicts a picture-sentence trial (P-S). Average BOLD signal for all subjects is plotted against the scan number. The coloured regions depict the mean BOLD signal +/- 1 standard deviation.}
  	\label{fig:boldroi}
\end{figure}

@@

From Table~\ref{tab:results} we can distinguish the fact that when PCA is used for feature extraction, the performance of GNB is lower than when other feature selection methods are used. The confidence interval of the predictive performance of GNB when PCA is used is 95\% $CI: 56.9 - 65.8$ which is significantly worse than applying GNB in combination with other feature selection methods. This suggests that maximizing the variance of the data does not maximize information. Another explanation could be that the data is not suited to be aligned by voxel index (see section~\ref{sec:dimensionalityReduction}). The performance listed in Table~\ref{tab:results} is for PCA using 149 components. This value was found by performing classification with a varying amount of components, the results for which are listed in figure~\ref{fig:pca_components}.

\begin{figure}
	\centering
	\includegraphics[width=90mm]{figures/pca_plot_by_index.png}
  	\caption{Classification accuracy of GNB after performing PCA with an increasing number of components.}
  	\label{fig:pca_components}
\end{figure}

@@

In this paper we have replicated earlier findings of decoding the cognitive states of looking at sentences or pictures from fMRI data. Specifically, by observing the patterns of activity for a group of subjects, we are able to predict with above-chance accuracy what the cognitive state of an unseen subject is. This shows that machine learning algorithms can be employed to capture useful information from noisy data, and generalise this information between subjects.\\
\indent We found that even though all of our classification procedures yielded well above chance-level classification performance, the differences between the classifiers was very seldom significant. Still, we consistently found that the Multivariate Gaussian (MVG) classifier outperformed the Gaussian Naive Bayes (GNB) classifier and Linear Support Vector machine (LSV). This finding suggests that the interactions between different brain areas contain information about the cognitive state, and that these functional connectivity generalises to an extent to different subjects.\\
\indent Still, the uncertainty in our predictions are relatively high, with maximum classification performance reaching close to 80\% accuracy for only two classes. The noisy nature of the fMRI data therefore makes it very difficult to draw hard conclusions when decoding a single cognitive state from a brain scan. This is something to be considered before employing fMRI as a diagnostic tool, or as evidence in court.\\
\indent Overall we can say that machine learning provides a feasible way of extracting useful information from fMRI data. Furthermore, it avoids a strict multiple comparisons correction that is required in null hypothesis testing by using cross-validation. Also, the identification of patterns of activity associated with tasks is encouraged, rather than attempting to identifying brain regions that are involved in tasks. In doing this, much can be learned about how information is represented and processed in the brain.

>>>>

\section{Methods}
\label{sec:methods}
We now continue by discussing the methods applied in our research. The general approach taken for each of the algorithms discussed consists of the following steps:
\begin{enumerate}
  \item Preprocessing
  \item Dimensionality reduction
  \item Training
  \item Testing
\end{enumerate}
Preprocessing was done to transform the dataset into a format suitable for input into the classification algorithm. Dimensionality reduction was required to cope with the high dimensionality of the dataset and may consist of both feature extraction as well as feature selection. Finally, the approach taken to training and testing the classifier deals with preventing overfitting the classifier to training data and verifying accuracy for cross-subject classification. The choices made in each of these steps may affect classifier accuracy, so to obain a complete understanding of classification accuracies we have tried several methods at each step. In the following sections, these methods will be described in more detail.\\
The implementation of the algorithms was done using Python combined with modules tailored to mathematics (NumPy, SciPy) and machine learning (ScikitLearn). The original dataset was made available as a MatLab file and we used preprocessing in MatLab to create a number of files containing modified versions of this data, such as normalization and ROI grouping (see section~\ref{ssec:preprocessing}). The source code for the algorithms is available from the authors on request.

@@

\textbf{\emph{ROI mapping.}} ROI mapping is a form of feature selection, wherein only a subset of the N-dimensional input space is used. As described in the introduction, ROIs are anatomically defined brain areas classified as relevant by an expert in the field. We used 7 of the 25 defined ROIs, labeled  'CALC' 'LIPL' 'LT' 'LTRIA' 'LOPER' 'LIPS' and  'LDLPFC'. These ROIs have previously been found to be most relevant in discriminating between the cognitive states of looking at sentences or pictures \cite{wa:2003betw}. To further reduce the number of features, average voxel values within these regions are taken as new input features. As with preprocessing, other higher-order features such as the standard deviation, kurtosis and skewness within ROIs did not increase predictability and were not included in the subsequent analysis. The disadvantage of using ROIs is that they require \textit{a priori} knowledge. When this knowledge is inaccurate, classification accuracy may not be optimal.

@@

\textbf{\emph{Temporal dimensionality reduction.}} The BOLD response peaks at approximately 5 seconds after a stimulus is presented (see Fig. 1). This means that the first scans after a stimulus may not represent activity related to this stimulus and should not be used for classification. To examine which scans prWindows API error 87: The parameter is incorrect.ovide maximal class information, we plotted the average BOLD activity per scan for all trials of all subjects, given a trial type and ROI. As can be seen in Figure~\ref{fig:boldroi}, scans 10 to 20 provide maximal information for the first stimulus, while scans 23 to 33 provide maximal information for the second stimulus. The other scans were excluded.      
\begin{figure}
	\centering
	\includegraphics[width=90mm]{figures/scan_hemodynamic_BOLD_roi}
  	\caption{Haemodynamic BOLD response for ROI "CALC". The red line depicts a sentence-picture trial (S-P) and the blue line depicts a picture-sentence trial (P-S). Average Bold activity for all subjects is plotted against the scan number. The coloured regions depict an interval of 1 standard deviation.}
  	\label{fig:boldroi}
\end{figure}

@@

From Table~\ref{tab:results} we can distinguish the fact that when PCA is used for feature extraction, the performance of GNB is lower than when other feature selection methods are used. The confidence interval of the predictive performance of GNB when PCA is used is 95\% $CI: 56.9 - 65.8$ which is significantly worse than applying GNB in combination with the other methods. This suggests that maximizing the variance of the data does not maximize information. Another explanation could be that the data is not suited to be aligned by voxel index (see section~\ref{sec:dimensionalityReduction}). The performance listed in Table~\ref{tab:results} is for PCA using 149 components. This value was found by performing classification with a varying amount of components, the results for which are listed in figure~\ref{fig:pca_components}.

\begin{figure}
	\centering
	\includegraphics[width=90mm]{figures/pca_plot_by_index.png}
  	\caption{Classification accuracy of GNB after performing PCA with a varying amount of components.}
  	\label{fig:pca_components}
\end{figure}

@@

In this paper we have replicated earlier findings of decoding the cognitive states of looking at sentences or pictures from fMRI data. Specifically, by observing the patterns of activity for a group of subjects, we are able to predict with above-chance accuracy what the cognitive state of an unseen subject is. This shows that machine learning algorithms can be employed to capture useful information from noisy data, and generalise this information between subjects.\\
\indent We found that even though all of our classification procedures yielded well above chance-level classification performance, the difference between the classifiers was seldomly significant. Still, we consistently found that the Multivariate Gaussian (MVG) classifier outperformed the Gaussian Naive Bayes (GNB) classifier and Linear Support Vector Machine (LSVM). This finding suggests that the interactions between different brain areas contain information about the cognitive state, and that these functional connectivity generalises to an extent to different subjects.\\
\indent Still, the uncertainty in our predictions are relatively high, with maximum classification performance reaching close to 80\% accuracy for only two classes. The noisy nature of the fMRI data therefore makes it very difficult to draw hard conclusions when decoding a single cognitive state from a brain scan. This is something to be considered before employing fMRI as a diagnostic tool, or as evidence in court.\\
\indent Overall we can say that machine learning provides a feasible way of extracting useful information from fMRI data. Furthermore, it avoids a strict multiple comparisons correction that is required in null hypothesis testing by using cross-validation. Also, the identification of patterns of activity associated with tasks is encouraged, rather than attempting to identifying brain regions that are involved in tasks. In doing this, much can be learned about how information is represented and processed in the brain.
